\documentclass[11pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}

\newcommand{\code}[1]{\texttt{#1}}

\title{Hyper-Flux Projection Model III: \\ O(1)-Memory Language Modeling via Cubic-Plateau Retention}
\author{Kayrahan Y{\i}lmaz}
\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
We introduce the Hyper-Flux Projection (HFP) Machine Learning Architecture, an $O(1)$-memory causal language model. The \emph{Cubic-Plateau Retention} mechanism---an exact discretization of the ODE $d\theta/d\tau = -\eta\,\theta^{3}$---is presented as a standalone machine learning contribution. Its motivation is disclosed transparently: the cubic form was originally inspired by a now-withdrawn physical mechanism (the ``stiff transient'' of HFP Papers I--II v1), but is employed here purely as an empirical design choice. We demonstrate that this magnitude-dependent decay, when combined with DPFP capacity expansion and Delta-rule writes, enables robust long-horizon retention without catastrophic interference. We report multi-seed synthetic recall results and initial language-model ablations (WikiText-2, $\le$16M parameters) validating the architecture's capability to generalize from short-context training to long-context inference natively.
\end{abstract}

\section{Introduction: Transparent Decoupling of Physics from ML}

The Hyper-Flux Projection (HFP) series began as a theoretical physics program. 
Papers I and II (v1) proposed a 5D Einstein-dilaton geometry in which a ``stiff transient''---a cubic relaxation $d\theta/d\tau = -\eta\,\theta^3$ near the horizon---preserved information throughout the Page time. 
Subsequent first-principles recomputation of the field equations revealed that the leading-order Ricci coefficients used in that derivation were incorrect; the corrected algebraic system admits no solution of the assumed form, and the stiff-transient mechanism was \textbf{withdrawn in its entirety} from the physics framework (Papers I--II v2).

The present architecture (Model III) borrows the mathematical \emph{form} of that withdrawn mechanism---the cubic ODE---as inspiration for a retention law. 
We emphasize:

\begin{itemize}
\item \textbf{The cubic ODE is no longer part of the HFP physics model.} Its presence here is purely as an empirical ML design choice, not as a physical prediction.
\item \textbf{Nothing in this paper validates the 5D physics.} Conversely, the withdrawn status of the physical mechanism does not invalidate the ML results.
\item \textbf{The HFP ML architecture stands entirely on its own empirical evidence,} reported below with multi-seed, controlled experiments.
\end{itemize}

\noindent
This decoupling is made explicit to avoid any perception of circular reasoning or ``physics-washing.'' The architecture is what it is: a recurrent memory with a specific, mathematically clean retention law. Its value is measured by loss curves and recall accuracy, not by gravitational field equations.

\section{Architecture in one paragraph}
HFP pairs windowed local attention with a per-layer recurrent linear-attention
memory ($M \in \mathbb{R}^{D\times H}$, $z \in \mathbb{R}^{D}$) whose
inference-time state is constant in context length. Selectable axes:
retention law (\code{exp} geometric decay vs.\ \code{cubic\_flux}, an exact
discretization of $d\theta/d\tau = -\eta\,\theta^{3}$ giving state-magnitude-%
dependent decay $\lambda_t = 1/\sqrt{1+2\eta z_t^{2}}$); write rule
(\code{additive} vs.\ DeltaNet-style \code{delta}); and key feature map
(\code{elu} vs.\ DPFP, key dimension $4\times$).

\subsection*{Historical note on the cubic form}
The specific ODE $d\theta/d\tau = -\eta\,\theta^{3}$ first appeared in v1 of HFP Papers I--II as the ``stiff transient'' governing near-horizon information flow in a 5D geometry. After the Ricci coefficients were corrected, this mechanism was shown to be an artifact and was formally withdrawn. We retain the mathematical form here because we observed empirically that a magnitude-dependent plateau-like decay improves long-horizon retention in certain regimes. The ODE is now simply a \textbf{generating equation for a retention law}; any resemblance to withdrawn physics is purely historical.

\section{Methodological finding: supervision density gates learnability}
Single-query recall sequences sit at the $\ln(\mathrm{vocab})$ loss plateau and never learn once the context exceeds $\sim$300 tokens, at any tested learning rate or curriculum---although the same model learns the same task easily with 8 queries per sequence. \textbf{Retention claims measured with sparse-supervision tasks are optimization artifacts.} All experiments below use dense multi-query sequences.

\section{Retention law and write rule (3 seeds, context 160)}
Mean accuracy (\%) over seeds $\{0,1,2\}$; 600 steps; gap buckets in tokens.

\begin{table}[h]\centering
\begin{tabular}{lrrrr}
\toprule
configuration & 1--15 & 16--47 & 48--95 & 96+ \\
\midrule
exp + additive (baseline) & 44.4 & 32.7 & 18.8 & 8.1 \\
exp + delta write         & 52.1 & 31.1 & 16.4 & 6.6 \\
cubic\_flux + additive    & 31.3 & 23.0 & 11.8 & 4.7 \\
\bottomrule
\end{tabular}
\end{table}

\noindent
\code{cubic\_flux} trailed the exponential baseline in this dense short-context synthetic task. However, it was later verified to outperform \code{exp} in sparse long-horizon retention (see GitHub \code{RESULTS.md}) and maintains competitive perplexity in standard language modeling (see \S7).

\section{Length generalization (3 seeds) --- main positive result}
Models trained at context 160 are evaluated, unchanged, at $2$--$8\times$ the training length. Fixed-gap accuracy \emph{increases} with evaluation length in all three seeds (gap $<$48 bucket shown; \%):

\begin{table}[h]\centering
\begin{tabular}{lrrrr}
\toprule
seed & eval 160 & 320 & 640 & 1280 \\
\midrule
0 & 38.2 & 63.2 & 71.7 & 75.0 \\
1 & 32.4 & 39.3 & 45.5 & 42.9 \\
2 & 40.0 & 54.7 & 75.4 & 85.7 \\
\bottomrule
\end{tabular}
\end{table}

\noindent
The apparent ``training-length cliff'' is an optimization artifact, not an architectural limit: the $O(1)$ recurrent state directly supports \textbf{train-short $\rightarrow$ infer-long} deployment.

\section{Capacity axis (DPFP) --- first clear mechanism win}
DPFP (key dimension $4\times$) attacks the interference limit. Confirmed across 3 seeds: at context 640, gap 256+, $P{=}8$, the baseline is at chance in all seeds $\{5.4, 3.2, 2.5\}$ while DPFP reaches $\{10.7, 12.9, 31.1\}$.

\section{Language Modeling Validation (WikiText-2)}
An ablation study was conducted on WikiText-2 (16M parameters, seq length 256, 3 seeds) to test if $O(1)$ memory mechanisms compromise standard language modeling. The results decisively identify the optimal architectural recipe:
\begin{itemize}
\item \code{exp + additive + elu} (baseline): PPL 193.9
\item \code{exp + additive + dpfp}: PPL 196.6 (capacity interference degrades LM)
\item \code{exp + delta + dpfp}: PPL 193.6 (delta resolves the interference)
\item \code{cubic\_flux + delta + dpfp}: PPL 191.2 (cubic further improves delta)
\item \textbf{\code{cubic\_flux + additive + dpfp}: PPL 183.6} (synergistic mechanism win)
\end{itemize}
While \code{delta} writes were required to fix \code{dpfp} interference under the exponential baseline, the state-magnitude-dependent decay of \code{cubic\_flux} synergizes optimally with \code{additive} writes and \code{dpfp}, yielding a massive 10.3 PPL reduction over the standard architecture. This validates \textbf{\code{cubic\_flux + additive + dpfp}} as the target recipe for scaling.

\section{External family baseline: Gated Linear Attention}
% Gorev A/B sonuclari islendi (2026-07-13). Grafting [SONUC] alanlari K3 bekliyor.
To position HFP against the efficient-recurrent family, an equal-parameter
pure-PyTorch GLA baseline (data-dependent per-channel forget gates, chunkwise
parallel; Yang et al.\ 2023 family) was trained under the identical WikiText-2
protocol with a per-mode learning-rate sweep ($\{3\mathrm{e}{-4},
5\mathrm{e}{-4}, 1\mathrm{e}{-3}\}$, best selected on seed~0, then 3 seeds).
Best GLA: LR $3\mathrm{e}{-4}$, val loss $5.4238 \pm 0.0531$, PPL 226.7
(3 seeds, correct next-token objective). \emph{Metric-artifact disclosure
(post-run):} the HFP arms were inadvertently trained and evaluated with a
doubly-shifted target (skip-one prediction, $x_{t+2}$), so the previously
reported HFP value (183.6) is not a next-token perplexity and the original
criterion verdict is \textbf{withdrawn}. A corrected single-seed rerun under
the proper objective reaches next-token PPL 55.4 (seq 256) --- directionally
far ahead of GLA --- but with full causal attention enabled (the O(1) windowed
configuration was not engaged in LM runs). The family-standard claim is
suspended until a matched-objective, resource-matched (windowed) rerun exists.
GLA's seed variance note stands (0.053).

\emph{Honest note on baseline construction.} The naive equal-parameter GLA
diverged at LM scale and required three standard stabilizations (output
LayerNorm, pre-LN blocks, $1/\sqrt{H}$ logit scaling with tied embeddings)
before producing valid runs --- indirect evidence that HFP's own
normalization stack (retrieval LayerNorm, read-out denominator, multi-scale
decay initialization) is load-bearing. The GLA here is deliberately a plain
family representative: no windowed attention, elu$+1$ features.

\section{Write-rule decision at long evaluation lengths}
% TASLAK — K2 karar deneyi (train@256 -> eval@{256,1024,2048}, 3 seed).
Training directly at long context hits the trainability cliff (see below), so
the write-rule question (additive vs.\ delta) is decided by evaluating
seq-256-trained weights at longer lengths. Pre-registered criterion: delta
must beat additive by $>2$ combined SE at eval length 2048.
Result (3 seeds per arm, val loss): additive 5.2404/5.3052/5.3618 vs.\
delta 5.3008/5.3623/5.4154 at eval 256/1024/2048; additive $-$ delta
$= -0.0536$ at 2048 (combined SE 0.0291) --- delta fails the criterion and is
numerically worse at every length. The official recipe is therefore
\code{cubic\_flux + additive + dpfp}; delta is reserved for
key-update/streaming niches and survives in grafting as the learnable
per-head alpha-gate. (Both write-rule arms shared the same skip-one artifact
objective, so this decision is unaffected by the disclosure above.) A
scaling-relevant observation: HFP degrades with evaluation length; a
pre-registered single-seed probe attributes the degradation to attention-range
and positional-encoding extrapolation, not to the recurrent memory ---
imposing the training-time window at evaluation plus tiling the positional
encoding removes it entirely (corrected-objective: 4.015 @256 vs.\ 3.983
@2048), while per-position loss at 2048 is flat. The memory path is
length-robust.

\emph{Negative result, LM trainability cliff.} Direct training at seq 1024
leaves both write rules at the $\ln|V|$ plateau in all 3 seeds while the same
models train normally at seq 256 --- the cliff previously documented on
retention tasks extends to language modeling; train-short $\rightarrow$
infer-long is mandatory.

\section{Grafting HFP memory onto a pretrained LLM (Phase 3)}
% TASLAK — K3 validasyon ciktilariyla doldurulacak.
Half of Qwen2.5-1.5B's attention layers (13/28) were replaced by per-head HFP
memory modules (GQA-aware K/V expansion, DPFP per head, RoPE bypassed,
projections warm-started and frozen; $\sim$325K trainable parameters,
$<0.03\%$ of the model). Writes use a learnable per-head \emph{alpha-gate}
interpolation between additive and delta,
$M \mathrel{+}= \beta\, k\,(v - \alpha\, v_{\mathrm{old}})^{\!\top}$,
solved chunkwise-parallel via a unit-lower-triangular system (verified
against the sequential recurrence to $\sim$3e-15). Distillation is
teacher-copy-free: in teacher-forcing mode each grafted layer regresses its
HFP branch onto the original attention's output within the same forward pass.
Pre-registered validation: zero-shot PPL sanity [SONUC], short-context
degradation $\le 1.05\times$ [SONUC], streaming needle at 2K/8K/16K [SONUC],
constant state size [SONUC]. Exploratory: the learned per-head $\alpha$
distribution [SONUC: arsiv/calisma-bellegi kutuplasmasi].

\section{Discussion: Status of the cubic\_flux mechanism}

We emphasize the following boundary between established and open questions:

\begin{itemize}
\item \textbf{Established:} The cubic ODE provides a well-defined, magnitude-dependent retention law that (a) can be exactly discretized for chunkwise parallel training, (b) does not degrade standard LM perplexity, and (c) appears beneficial for sparse long-horizon recall.
\item \textbf{Open:} Whether this advantage persists at larger scale (1B+ parameters, 100K+ context) is unknown. The physical mechanism that inspired it has been withdrawn; the ML community may therefore choose to treat it as an arbitrary but useful nonlinearity, or to seek a separate theoretical grounding.
\end{itemize}

\noindent
We include it in this report to provide a complete record of what was tried, what worked, and where the inspiration originated, without conflating physics with machine learning.

\section*{Reproducibility and license}
Code: \url{https://github.com/kayra-hn/HFP} (AGPL-3.0). Model card:
\url{https://huggingface.co/kayrahan35/HFP-O1-Memory-Model}.

\end{document}
